
```{r setup, include=FALSE}

```

---
title: 'Portofolio 5: Final Project: Tracing the Enlightenment Thought in the Encyclopédie'
author: 'Kristian Jensby'
date: 'created on 20 May 2026 and updated `r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---


### Loading packages
```{r packages}

library(tidyverse)
library(here)
library(pdftools)
library(tidytext)
library(textdata) 
library(ggwordcloud)
```


### Importing the Encyclopedia text by D'Alembert and Diderot:
```{r get-document}
Encyclopedia_path <- here("data","Diderot_Encyclopedia.pdf")
Encyclopedia_text <- pdf_text(Encyclopedia_path)
```

### In order do to a sentiment analysis I of the Encyclopedia I need to structurise the text by doing some wrangling:

- Split up pages into separate lines (separated by `\n`) using `stringr::str_split()`
- Unnesting into regular columns using `tidyr::unnest()`
- Removing leading/trailing white space with `stringr::str_trim()`

```{r split-lines}
Encyclopedia_df <- data.frame(Encyclopedia_text) %>% 
  mutate(text_full = str_split(Encyclopedia_text, pattern = '\n')) %>% 
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full)) 

```

Now each line, on each page, is its own row, with extra starting & trailing spaces removed. 

### Getting the tokens (individual words) in tidy format

I am using `tidytext::unnest_tokens()` (which is pulling from the `tokenizer`) package, to split columns into tokens. I am interested in *words*, so that's the token I'll use:

```{r tokenize}
Encyclopedia_tokens <- Encyclopedia_df %>% 
  unnest_tokens(word, text_full)

# Now each word has its own row
```

Counting the words
```{r count-words}
Encyclopedia_wc <- Encyclopedia_tokens %>% 
  count(word) %>% 
  arrange(-n)
Encyclopedia_wc
```

Now I am noticing that a lot of words  show up frequently that I am not so interested in for this analysis (e.g. "the", "of", "and", etc.). These are  *stop words* and now I will remove them.

### Removing stop words:

I use `?stop_words` and `View(stop_words)`to look at documentation for stop words lexicons.

I will *remove* stop words using `tidyr::anti_join()`:
```{r stopwords}
Encyclopedia_stop <- Encyclopedia_tokens %>% 
  anti_join(stop_words) %>% 
  select(-Encyclopedia_text)
```

Now I check the counts again: 
```{r count-words2}
Encyclopedia_swc <- Encyclopedia_stop %>% 
  count(word) %>% 
  arrange(-n)
```

### Creating a wordcloud of Encyclopedia report words for a thematic overview 


```{r wordcloud-prep}



# Filtering words into the most 80 frequent as the wordcloud as it is a simple thematic overview
Encyclopedia_top80 <- Encyclopedia_stop %>% 
  count(word) %>% 
  arrange(-n) %>% 
  head(80)
Encyclopedia_top80
```

```{r wordcloud}
Encyclopedia_cloud <- ggplot(data = Encyclopedia_top80, aes(label = word)) +
  geom_text_wordcloud() +
  theme_minimal()

Encyclopedia_cloud
```

Now I am customising it a bit for the sake of the visualisation:
```{r wordcloud-pro}
ggplot(data = Encyclopedia_top80, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "square") +
  scale_size_area(max_size = 16) +
  scale_color_gradientn(colors = c("blue","white","red")) +
  theme_minimal()
```

### Sentiment analysis

```{r afinn}

# Downloading the afiin lexicon
get_sentiments(lexicon = "afinn")

```

```{r nrc}
#Downloading the nrc lexicon
get_sentiments(lexicon = "nrc")

```

Now I am doing sentiment analysis on the Encyclopedia text data using afinn and nrc. 


### Sentiment analysis with afinn: 

First, I bind words in `Encyclopedia_stop` to `afinn` lexicon:
```{r bind-afinn}
Encyclopedia_afinn <- Encyclopedia_stop %>% 
  inner_join(get_sentiments("afinn"))
Encyclopedia_afinn
```

Finding some counts (by sentiment ranking):
```{r count-afinn}
Encyclopedia_afinn_hist <- Encyclopedia_afinn %>% 
  count(value)

# Plotting them: 
ggplot(data = Encyclopedia_afinn_hist, aes(x = value, y = n)) +
  geom_col(aes(fill = value)) +
  theme_bw()
```

Summarisiing sentiment for the text: 
```{r summarize-afinn}
Encyclopedia_summary <- Encyclopedia_afinn %>% 
  summarize(
    mean_score = mean(value),
    median_score = median(value)
  )
Encyclopedia_summary 
```

The mean and median indicate a slightly positive overall sentiments based on the AFINN lexicon. 

### NRC lexicon for sentiment analysis

 As above, I'll use inner_join() to combine the Encyclopedia non-stopword text with the nrc lexicon: 

```{r bind-bing}
Encyclopedia_nrc <- Encyclopedia_stop %>% 
  inner_join(get_sentiments("nrc"))
```

Then I check which words are being excluded by using `anti_join()`:

```{r check-exclusions}
Encyclopedia_exclude <- Encyclopedia_stop %>% 
  anti_join(get_sentiments("nrc"))

# View(Encyclopedia_exclude)

# Counting to find the most excluded:
Encyclopedia_exclude_n <- Encyclopedia_exclude %>% 
  count(word, sort = TRUE)

head(Encyclopedia_exclude_n)
```

Now I find some counts: 
```{r count-bing}
Encyclopedia_nrc_n <- Encyclopedia_nrc %>% 
  count(sentiment, sort = TRUE)

# And I plot them:

ggplot(data = Encyclopedia_nrc_n, aes(x = sentiment, y = n)) +
  geom_col(aes(fill = sentiment))+
  theme_bw()
  

```

Or I can count by sentiment and word, then facet:
```{r count-nrc}
Encyclopedia_nrc_n5 <- Encyclopedia_nrc %>% 
  count(word,sentiment, sort = TRUE) %>% 
  group_by(sentiment) %>% 
  top_n(5) %>% 
  ungroup()

Encyclopedia_nrc_gg <- ggplot(data = Encyclopedia_nrc_n5, aes(x = reorder(word,n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, ncol = 2, scales = "free") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Word", y = "count")

# Showing it
Encyclopedia_nrc_gg

# Saving it
ggsave(plot = Encyclopedia_nrc_gg, 
       here("figures","Encyclopedia_nrc_sentiment.png"), 
       height = 12, 
       width = 7)
        labs(x = "Word", y = "count")

```

