---
title: "Assignment-3"
author: "Kristian Jensby"
output: html_document
date: "2026-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: 'Portofolio 3: Text mining, sentiment analysis, and visualization in GoT '
date: 'created on 26 March 2026 and updated `r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---



```{r packages}

library(tidyverse)
library(here)
library(pdftools)
library(tidytext)
library(textdata) 
library(ggwordcloud)
```


### Getting the GOT Book:
```{r get-document}
got_path <- here("data","got.pdf")
got_text <- pdf_text(got_path)
```

Example: Getting text from a single page (e.g. Page 9)? 
```{r single-page}
got_p9 <- got_text[9]
got_p9
```

### Wrangling:

- Splitting up pages into separate lines (separated by `\n`) using `stringr::str_split()`
- Unnesting into regular columns using `tidyr::unnest()`
- Removing leading/trailing white space with `stringr::str_trim()`

```{r split-lines}
got_df <- data.frame(got_text) %>% 
  mutate(text_full = str_split(got_text, pattern = '\n')) %>% 
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full)) 

```

Now each line, on each page, is its own row, with extra starting & trailing spaces removed. 

### Get the tokens (individual words) in tidy format

Using `tidytext::unnest_tokens()` (which pulls from the `tokenizer`) package, to split columns into tokens. I am interested in *words*, so that's the token I'll use:

```{r tokenize}
got_tokens <- got_df %>% 
  unnest_tokens(word, text_full)

# Now each word has its own row
```

Counting the words
```{r count-words}
got_wc <- got_tokens %>% 
  count(word) %>% 
  arrange(-n)
got_wc
```

Now I am noticing a lot of words  show up frequently that I am not so interested in for this analysis (e.g. "the", "and", "to", etc.). These are  *stop words* and now I will remove them. 

### Removing stop words:

I use `?stop_words` and `View(stop_words)`to look at documentation for stop words lexicons.

I will *remove* stop words using `tidyr::anti_join()`:
```{r stopwords}
got_stop <- got_tokens %>% 
  anti_join(stop_words) %>% 
  select(-got_text)
```

Now I check the counts again: 
```{r count-words2}
got_swc <- got_stop %>% 
  count(word) %>% 
  arrange(-n)
```

Then I want to get rid of all the numbers (non-text) in `got_stop`.
```{r skip-numbers}

got_no_numeric <- got_stop %>% 
  filter(is.na(as.numeric(word)))
```

### Creating a word cloud of GoT report words (non-numeric)


```{r wordcloud-prep}

length(unique(got_no_numeric$word))

# Filtering words into the most 100 frequent
got_top100 <- got_no_numeric %>% 
  count(word) %>% 
  arrange(-n) %>% 
  head(100)
got_top100
```

```{r wordcloud}
got_cloud <- ggplot(data = got_top100, aes(label = word)) +
  geom_text_wordcloud() +
  theme_minimal()

got_cloud
```

Now I am customising it a bit:
```{r wordcloud-pro}
ggplot(data = got_top100, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "pentagon") +
  scale_size_area(max_size = 16) +
  scale_color_gradientn(colors = c("orange","red","pink")) +
  theme_minimal()
```

### Sentiment analysis

"afinn": Words ranked from -5 (very negative) to +5 (very positive)
```{r afinn}
get_sentiments(lexicon = "afinn")

# Looking at the pretty positive words:
afinn_pos <- get_sentiments("afinn") %>% 
  filter(value %in% c(3,4,5))

afinn_pos
```

bing: binary, "positive" or "negative"
```{r bing}
get_sentiments(lexicon = "bing")

```

Now nrc:
```{r nrc}
get_sentiments(lexicon = "nrc")

```

Now I am doing sentiment analysis on the GoT text data using afinn, and nrc. 


### Sentiment analysis with afinn: 

First, I bind words in `got_stop` to `afinn` lexicon:
```{r bind-afinn}
got_afinn <- got_stop %>% 
  inner_join(get_sentiments("afinn"))
got_afinn
```

Finding some counts (by sentiment ranking):
```{r count-afinn}
got_afinn_hist <- got_afinn %>% 
  count(value)

# Plotting them: 
ggplot(data = got_afinn_hist, aes(x = value, y = n)) +
  geom_col(aes(fill = value)) +
  theme_bw()
```

Investigating some of the words in a bit more depth:
```{r afinnneg5}
# '-5' words?
got_afinnneg5 <- got_afinn %>% 
  filter(value == -5)
```

```{r afinnneg5-more}
# Checking the unique negative 5-score words:
unique(got_afinnneg5$word)

# Counting & plotting them
got_afinnneg5 <- got_afinnneg5 %>% 
  count(word, sort = TRUE) %>% 
  mutate(word = fct_reorder(factor(word), n))


ggplot(data = got_afinnneg5, aes(x = word, y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw()

```

Summariziing sentiment for the text: 
```{r summarize-afinn}
got_summary <- got_afinn %>% 
  summarize(
    mean_score = mean(value),
    median_score = median(value)
  )
got_summary 
```

The mean and median indicate a slightly negative overall sentiments based on the AFINN lexicon. 

### NRC lexicon for sentiment analysis

 As above, I'll use inner_join() to combine the got non-stopword text with the nrc lexicon: 

```{r bind-bing}
got_nrc <- got_stop %>% 
  inner_join(get_sentiments("nrc"))
```

Then I check which words are being excluded by using `anti_join()`:

```{r check-exclusions}
got_exclude <- got_stop %>% 
  anti_join(get_sentiments("nrc"))

# View(got_exclude)

# Counting to find the most excluded:
got_exclude_n <- got_exclude %>% 
  count(word, sort = TRUE)

head(got_exclude_n)
```

Now I find some counts: 
```{r count-bing}
got_nrc_n <- got_nrc %>% 
  count(sentiment, sort = TRUE)

# And I plot them:

ggplot(data = got_nrc_n, aes(x = sentiment, y = n)) +
  geom_col(aes(fill = sentiment))+
  theme_bw()
```

Or I can count by sentiment and word, then facet:
```{r count-nrc}
got_nrc_n5 <- got_nrc %>% 
  count(word,sentiment, sort = TRUE) %>% 
  group_by(sentiment) %>% 
  top_n(5) %>% 
  ungroup()

got_nrc_gg <- ggplot(data = got_nrc_n5, aes(x = reorder(word,n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, ncol = 2, scales = "free") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Word", y = "count")

# Showing it
got_nrc_gg

# Saving it
ggsave(plot = got_nrc_gg, 
       here("figures","got_nrc_sentiment.png"), 
       height = 12, 
       width = 7)

```

Hereby I can show that "bastard" is showing up in NRC lexicon as the sentiments "disgust", "negative" and "sadness":
```{r nrc-bastard}
bastard <- get_sentiments(lexicon = "nrc") %>% 
  filter(word == "bastard")

bastard
```

## Explanation of the visual together with a question that it addresses
With this sentiment analysis of the first book of The Song of Ice and Fire or commonly known as the Game of Thrones, I am interested in the question regardering how and if it can me implicated that the langugage and vocabulary of this book is overly consisting of words with a negative and or a bit vulgar sentiment.
And referring to the visualisations of this portfolio it is highly evident that this is the case. E.g. the high occurences of words like "bastards", "bastard" and "bitch" etc. The further sentiment visualisation show that the sentiments like "negative", "anger", "sadness" and "disgust" have a high concentration of words in these sentimental categories.
the occurrence of words with really negative sentiments. 
However it is important to note that words like "lord" is sorted in the disgust sentiment, which for this analysis is relevant it is a word with not necessarily negative connotations.

