---
title: "Danish Kings"
author: "Kristian Jensby"
date: "2025-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the separator: 

A1: The separator is ";" (semicolon).

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
library(tidyverse)
library(dplyr)
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/dkkings.csv")
glimpse(kings1)
class(kings1)
ncol(kings1)

kings2 <- read_csv("data/dkkings.csv")
glimpse(kings2)
class(kings2)
ncol(kings2)

kings3 <- read.csv2("data/dkkings.csv")
glimpse(kings3)
class(kings3)
ncol(kings3)

kings4 <- read_csv2("data/dkkings.csv")
glimpse(kings4)
class(kings4)
ncol(kings4)

```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object.

A1: The 'tidyverse' functions are read_csv and read_csv2.

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

A2: When you run 'class()' on the 'kings' object created with a tidyverse function it should return: 
  "spec_tbl_df"
  "tbl_df"
  "tbl" 
  "data.frame" 
This should yield the information that the 'kings' object is a tibble.

3. How many columns does the object have when created with these different functions?

A3: read.csv and read_csv objects have 1 column.
    read.csv2 and read_csv2 object have 5 columns.

4. Show the dataset so that we can see how R interprets each column.

4A. I utilise the glimpse() function.
    read.csv2 interprets columns 'Born', 'Dead', 'Reign_start' and 'Reign_end' as the type <int> (integer) 
    read_csv2 interprets columns 'Born', 'Dead', 'Reign_start' and 'Reign_end' as the type <dbl> (double)


```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv2("data/dkkings.csv")

glimpse(kings)

class(kings)

ncol(kings)

head(kings)

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE
kings <- kings %>% 
  mutate(Duration = reign_end - reign_start)
glimpse(kings)
```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# base-R:
mean_duration <- mean(kings$Duration, na.rm = TRUE)
mean_duration

# I get a value of 20.2963.

# Tidyverse:
tidy_mean_duration <- kings %>%
  summarise(mean(Duration, na.rm = TRUE))
tidy_mean_duration

# I get a tibble containing the mean. 
```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

A: 26 kings enjoyed a longer-than-average duration of reign.

```{r} 
kings %>% 
  filter(Duration > mean_duration) %>% 
  select(name, Duration) %>% 
  count()

above_average_kings <- kings %>% 
  filter(Duration > mean_duration) %>% 
  select(name, Duration)
  view(above_average_kings)
```

## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
kings %>% 
  select(name, Duration) %>% 
  arrange(desc(Duration))

top_kings <- kings %>%
  arrange(desc(Duration)) %>%
  slice(1:3)

top_kings <- top_kings %>%
  mutate(Days = Duration * 365)

top_kings <- top_kings %>%
  mutate(
    Leap_years = Duration %/% 4,
    Days_accounting_for_leap_years = Duration * 365 + Leap_years)
glimpse(top_kings)
```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

A: The long-term trend in the duration of reign among Danish monarchs is that duration of reign increases through history. 

```{r}
kings %>% 
  mutate(Duration = reign_end - reign_start) %>% 
  mutate(Midyear = reign_end - Duration/2) %>% 
  ggplot(aes (x = Midyear, y = Duration)) +
  geom_point(col = "black") + 
  geom_smooth() + 
  theme_classic() +
  labs(title = " Duration of Danish Kings' Reigns over Time",
       x = "Time (years)",
       y = "Duration of Reign (years)")
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


